Exploring the NOAA Storm Database to uncover the events with biggest effect on Population Health and the Economy
=======================

Synopsis
---------------------------
describes and summarizes the data analysis in less than 10 sentences

The purpose of this report is to use the information stored in the  NOAA Storm Database to find out which types of events have the most severe effects on Population Health and the Economy. This is **important**, because if we are able to **stratify** these events with respect to the aforementioned categories, given the fact that there are **finite resources that can be dedicated for prevention and protection against these events**, we can focus these resources to target the events with the biggest impact, thus **maximizing the efficiency of those resources**.

We will start with an exploration of the data, in an attempt to find out what sort of information is included, which are the variables in the dataset we need to focus on, whether we need to clean or impute the data and in general any assumptions we need to have in mind for our analysis.

The questions we are looking to answer do not require statistical inference. As a starting point, descriptive statistics should provide us with enough understanding to answer. However, it might be meaningfull to retrospectively look for other underlying relations in our data.

After loading and processing the dataset, we will do any necessary transformations 
Data Processing
---------------
describes how the data were loaded into R and processed for analysis

```{r}
setwd("D:\\Coursera\\Reproducible Research\\RepData_PeerAssessment2")
library(knitr)
library(ggplot2)
library(dplyr)
```

```{r echo=TRUE, cache=TRUE}
data <- read.csv("data/repdata-data-StormData.csv.bz2")
data <-tbl_df(data)
str(data)
```

The data consist of 902297 of 37 variables. After consulting the [Storm Data documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "Storm Data documentation") and the [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf "FAQ") documents we decided to focus on a specific subset of variables namely:
FATALITIES, INJURIES, PROPDMG, and CROPDMG . We will create a subset dataset called df that only contains these variables. 

```{r}
df <- select(data, EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)
```
We then proceed to aggregate by event type.
```{r}
df_by_event <- group_by(df, EVTYPE)
```

Results
=======
the main results are presented

Types of events that are most harmful with respect to population health
------------------------------------------------------------------------
Here we can see the Fatalities and Injuries by event, ordered by most Fatalities and then by most Injuries (top ten).
```{r}
pop_health <- summarise(df_by_event, Fatalities = sum(FATALITIES), Injuries = sum(INJURIES))%.%arrange(desc(Fatalities),desc(Injuries))
(pop_health <- pop_health[1:10,])
```

```{r}
g1 <- ggplot(pop_health, aes(x=EVTYPE,y=Fatalities,fill=Injuries))+geom_bar(stat="identity")
g1 + theme_bw() + ggtitle("Fatalities and injuriesby event type")
```

Types of events with the greatest economic consequences
-------------------------------------------------------
Here we present the Economic effects ordered by Property damage and then by Crop damage (top ten)
```{r}
economy <- summarise(df_by_event, Property = sum(PROPDMG), Crop = sum(CROPDMG))%.%arrange(desc(Property),desc(Crop))
(economy <- economy[1:10,])
```

```{r}
g2 <- ggplot(economy, aes(x=EVTYPE, y=  Property, fill = Crop)) +geom_bar(stat="identity")
g2 + theme_bw() + ggtitle ("Economic effects by event type")
```

Conclusions
===========
